Accounting for fine-scale correlation structures in hidden Markov models

### 2020-07-12 Report/Sections/appendix.tex
Update appendix - before moving tables and figures

\section*{Appendix}

\setcounter{subsection}{0}

\subsection{Detailed Description of Data Simulation}

$\hat{Y}^*_{t,s^*}$ was simulated using the following procedure. Note that the $k^{th}$ Fourier mode of $\hat{Y}^*_{t,s^*}$ is denoted as $\hat{Y}^{*(k)}_{t,s^*}$:

\begin{align*}
	(\hat{Y}^{*(0)}_{t,0}|X^*_{t,0} = i) &\sim \mathcal{N} \left(0, \sigma^{*(i)} \right) & \\
	%
	(\hat{Y}^{*(0)}_{t,s^*}|X^*_{t,s^*} = i) &\sim \mathcal{N} \left(\phi^{*(i)} * \hat{Y}^{*(0)}_{t,s^*-1}, \sigma^{*(i)} \right), & s^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \\
	%
	\hat{Y}^{*(k)}_{t,s^*} &= a_{t,s^*}^{(k)} i\sqrt{b^{(k)}_{t,s^*}}, & k = 1,\ldots,49 \\
\end{align*}
\begin{align*}
    a_{t,s^*}^{(k)} &\sim  \left\{\begin{array}{lr}
	-1 & \quad w.p. \enspace 1/2 \\
	1  & \quad w.p. \enspace 1/2
	\end{array}\right. \\
	%
	(b^{(k)}_{t,s^*}|X^*_{t,s^*}  = 1) &\sim {\rm{Gamma}}(1/k^2, 1) \\
	%
	(b^{(k)}_{t,s^*}|X^*_{t,s^*} = 2) &\sim \left\{\begin{array}{lr}
	{\rm{Gamma}}(1/k^2, 1), \quad & k \notin \{1,2\} \\
	{\rm{Gamma}}(100,1), & k = 1 \\
	{\rm{Gamma}}(50,1), & k = 2
	\end{array}\right. 
\end{align*}

\begin{align*}
    \hat{Y}^{*(50)}_{t,s^*} &= 0 & \\
	%
	\hat{Y}^{*(k)}_{t,s^*}  &= -\hat{Y}^{*(100-n)}_{t,s^*}, & \qquad k = 51,\ldots,99
\end{align*}

$Y^*_{t,(100s^*+1):(100s^*+100)}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,s^*}$:

$$Y^*_{t,(100s^*+1):(100s^*+100)} = IDFT\left(\hat{Y}^*_{t,s^*}\right), \qquad s^* = 0,\ldots,\lfloor Y_t/2 \rfloor - 1$$

There are several practical reasons behind this construction. First, $\hat{Y}^*_{t,s^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,s^*}$ so that its inverse Fourier transform is real-valued. $\hat{Y}^{*(k)}_{t,s^*}$ also decays like $1/k^2$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window. Note, however, that $Y^*_{t,s^*}$ is not continuous \textit{between} windows, as can be seen in (fig \ref{fig:fourier_example}). However, the jumps are not too severe since $\hat{Y}^{*(0)}_{t,s^*}$ and $\hat{Y}^{*(0)}_{t,s^*+1}$ are highly correlated. See (fig. \ref{fig:sim_data}) for details.

From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$ after setting $\tilde{f}$. We pick $\tilde{f} = 10$ periods per window, or 5 hertz. Note that $Z^{*(1)}_{t,s^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,s^*}\right)$, so selecting $\sigma^{*(i)}$ and $\phi^{*(1)}$ appropriately will ensure correct distribution.

Note that $Z^{*(2)}$, is the sum of gamma distributions with the same scale parameter, so the distribution of $Z^{*(2)}$ is also a Gamma distribution:

$$Z^{*(2)} = \sum_{k=1}^{10} b^{(k)}_{t,s^*}$$
$$Z^{*(2)} \sim {\rm{Gamma}}$$

@@

First, we simply use down-sampling and only record every $w^{th}$ time step of $\hat{Y}_t^*$. This both reduces the space of $\hat{Y}_t^*$ to $\mathbb{C}^{\lfloor S^*_t / w \rfloor \times w}$ and ensures that none of the sliding windows overlap when taking the STFT. This is important because HMMs assume temporal independence between observations. Next, the dimension of $\hat{Y}_t^*$ can be cut in half by recognizing that the $Y_t^*$ is real-valued, and therefore $\hat{Y}_{t,k}^*$ is equal to the complex-congugate of $\hat{Y}_{t,w-k-1}^*$. Finally, by Parseval's Thereom we have that:

>>>>

\begin{proof}{Proof of Theorem 1}{}
Combining equation (\ref{eqn:CarHMM_to_OU}) with equation (\ref{eqn:OU_sol}) gives:
%
\begin{align*}
\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i^* \right) \sim \mathcal{N}\left((1-\phi^{(i^*)}) \mu^{(i^*)} + \phi^{(i^*)} Y_{(s-1) \Delta t}, \enspace \sigma^{2(i^*)} \right),\\
s = 1, \ldots, T-1
\end{align*}
%
If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t: (s-1) \Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:

\begin{align*}
\left(Y'_s| X'_s = i^* \right) \sim \mathcal{N}\left((1-\phi^{(i^*)}) \mu^{(i^*)} + \phi^{(i^*)} Y'_{s-1}, \enspace \sigma^{2(i^*)} \right)\\
s = 2, \ldots, T
\end{align*}
%
$X'$ follows a Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s+1})$ is consistent with that of a CarHMM with normal emission distributions.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{proof}{Description of simulated data}{}

$\hat{Y}^*_{t,t^*}$ was simulated using the following procedure. The $k^{th}$ Fourier mode of $\hat{Y}^*_{t,t^*}$ is denoted as $\hat{Y}^{*(k)}_{t,t^*}$:

\begin{align}
	(\hat{Y}^{*(0)}_{t,0}|X^*_{t,0} = i^*) &\sim \mathcal{N} \left(0, \sigma^{*(i^*)} \right) & \nonumber \\
	%
	(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*) &\sim \mathcal{N} \left(\phi^{*(i^*)} * \hat{Y}^{*(0)}_{t,t^*-1}, \sigma^{*(i^*)} \right), & t^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \label{eqn:y_hat0} \\
	%
	\hat{Y}^{*(k)}_{t,t^*} &= a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, & k = 1,\ldots,49 \nonumber
\end{align}

\begin{align*}
    a_{t,t^*}^{(k)} &\sim  \left\{\begin{array}{lr}
	-1 & \quad w.p. \enspace 1/2 \\
	1  & \quad w.p. \enspace 1/2
	\end{array}\right. \\
	%
	(b^{(k)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
	%
	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
	{\rm{Gamma}}(5/k^2, 1), \quad & k \notin \{1,2\} \\
	{\rm{Gamma}}(250,1), & k = 1 \\
	{\rm{Gamma}}(50,1), & k = 2
	\end{array}\right. 
\end{align*}

\begin{align*}
    \hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
	%
	\hat{Y}^{*(k)}_{t,t^*}  &= -\hat{Y}^{*(100-k)}_{t,t^*}, & \qquad k = 51,\ldots,99
\end{align*}

$Y^*_{t,t^*,1:100}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:

$$Y^*_{t,t^*,1:100} = IDFT\left(\hat{Y}^*_{t,t^*}\right), \qquad t^* = 1,\ldots,\lfloor Y_t/2 \rfloor$$

$\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued. $\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window. $Y^*_{t,t^*}$ is not continuous \textit{between} windows, but the jump discontinuities are not very severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+1}$ are highly correlated. See (fig. \ref{fig:sim_data}) for details.

From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$. We pick $\tilde{\omega} = 10$ periods per window, or 5 hertz. To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use (eqn \ref{eqn:y_hat0}):
%
$$\left(Z^{*(1)}_{t,t^*} | X^*_{t,t^*} = i^* \right) = \left(\mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right) | X^*_{t,t^*} = i^* \right) \sim \mathcal{N} \left(\phi^{*(i^*)} * Z^{*(1)}_{t,t^*-1}, \sigma^{*(i^*)} \right)$$
%
$Z^{*(2)}_{t,t^*}$ is the sum of Gamma-distributed random variables with the same scale parameter, so the distribution of $Z^{*(2)}_{t,t^*}$ is also a Gamma distribution:
%
$$Z^{*(2)}_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right)$$
\end{proof}

\begin{proof}{Likelihood of Simulation study model}{}

The overall likelihood of the CarHHMM-DFT model is as follows:
%
$$\calL_{\text{CarHHMM-DFT}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
%
where:
%
\begin{align*}
P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
&f^{(N)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
\end{align*}
%
$f^{(i)}(y_t;\Theta^{(i)})$ is the emission distribution of the dive duration $y_t$ conditioned on the fact that $X_t = i$.

$\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale chain:
%
$$\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(i)},\Gamma^{*(i)}\right) = \delta^{*(i)} \prod_{t=2}^T \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i^*)}) \mathbf{1}_N$$
%
where $P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i)})$ is an $N^* \times N^*$ diagonal matrix with $i^*i^*$th entry equal to $f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$.
%
$f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$ is the emission distribution of $Z^*_{t,t^*}$ conditioned on the fact that $X^*_{t,t^*} = i^*$ and $Z^*_{t,t^*-1}$.

\end{proof}

@@

First, we simply use down-sampling and only record every $w^{th}$ time step of $\hat{Y}_t^*$. This both reduces the space of $\hat{Y}_t^*$ to $\mathbb{C}^{\lfloor t^*_t / w \rfloor \times w}$ and ensures that none of the sliding windows overlap when taking the STFT. This is important because HMMs assume temporal independence between observations. Next, the dimension of $\hat{Y}_t^*$ can be cut in half by recognizing that the $Y_t^*$ is real-valued, and therefore $\hat{Y}_{t,k}^*$ is equal to the complex-congugate of $\hat{Y}_{t,w-k-1}^*$. Finally, by Parseval's Thereom we have that:

### 2020-07-12 Report/Sections/intro.tex
Update appendix - before moving tables and figures

At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}. One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.

>>>>

All HMMs presented so far assume Markovian dynamics in the underlying process, i.e. that any observation $Y_t$ depends only on the behavioral state $X_t$ and $Y_{t-1}$ when conditioned on all previous time steps. At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.  With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine-scale processes are likely to persist.

One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
